---
title: "ariadne: Integrate Knowledge in your Bioinformatic Analysis"
author:
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
# bibliography: ../inst/REFERENCES.bib
date: "`r doc_date()`"
package: "`r pkg_ver('ariadne')`"
vignette: >
  %\VignetteIndexEntry{ariadne: Integrate Knowledge into your Bioinformatic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

All domains of biology have to do with relational information. From genes to
organisms, each and every biological entity has some link to other entities. For
example, bugs are friends or foes with other bugs, have foods they like and
foods they hate (sometimes to death). All these links we call relational
information, which becomes knowledge once it is smartly combined with other
pieces and kept safe and open in a wonderfully organised place, usually a
database.

Building a database from scratch is no easy task. Fortunately, nowadays
databases for all kinds of relational information are available: UniProt for
proteins, Rhea for reactions, KEGG for pretty much anything, and many others,
which keep growing together with the field they support. As a scientist, having
access to such vast knowledge from past research can truly make us "stand on the
shoulders of giants".

In practice, though, it is not always easy (or even possible) to leverage such
knowledge in routine bioinformatic analyses due to three main challenges. First,
data is usually deposited somewhere online, so there must be a fast and reliable
method to retrieve and import it in the local environment. Second, data size
often exceeds the disk space of a typical laptop, which requires smart ways to
efficiently subset and store data in a lightweight, accessible format. Third,
unless you are lucky, data will need a tidy-up before it is compatible with your
dataset and other resources, necessitating a systematic approach to integrate
information from multiple resources.

## Motivation

Sometimes biology seems like a huge labyrinth with no simple paths between its
building blocks (genes, proteins, organisms, etc.). Even when the path to take
is clear, many obstacles may hide on the way: inaccessible resources, long query
times, incompatible formats, just to name a few. Crossing the labyrinth without
a strategy is a seriously boring and discouraging task.

Luckily, ariadne is here to help. By "Assembling Relational Information Across
the Database Network", ariadne provides a harmonised graph of interlinked
resources and a rich toolkit to navigate its underexplored paths. 

The harmonised graph of resources was built with a reproducible pipeline hosted
in the companion package `ariadne.db`, and it is openly available on [Zenodo](https://doi.org/10.5281/zenodo.18788725), including its previous
versions.


## Implementation

The ariadne jargon:

- **feature type**: Examples: ko (KEGG Orthologues), uniref90
- **level**: a unique element belonging to a feature type. The term _feature_
  could be used as well, but it is not ideal due to its many closely related
  meanings. Examples: 563 (NCBI id for E. coli), C00246 (KEGG compound id for
  butyrate).
  
- **linkmap**: 2-column data.frame where columns are the feature types and rows
  the links between them. This is how relational information is represented in
  ariadne. It is often named with the syntax `x2y`, where x and y are the names
  of the feature types in the first and second columns, respectively. Sometimes,
  there is also a third column with the names for the target feature type y.
- **resource**: a database from where linkmaps between feature types can be
  retrieved. Examples: KEGG, Rhea or UniProt.
- **weave**: the operation of sequentially linking feature types



Looking inside ariadne:

- The resource graph is represented as an `igraph` object, where edges are
  resources and nodes feature types. This is a natural choice for relational
  data and enables applying graph theory to select relevant paths from one
  feature type to another.
- Resources are queried via fast `httr2` requests or using pre-existing APIs.
  Whenever possible, even faster SPARQL queries are used to fetch linkmaps.
- Linkmaps are automatically cached with `BiocFileCache` and efficiently stored
  and imported using the Apache parquet columnar format from `arrow`.
- `MultiFactor` weaves the output linkmap under the hood.


## Applications

ariadne provides a rapid and robust approach to incorporate taxonomic
and functional module analyses into your bioinformatic workflow. Despite the
presence of specialised bioinformatic tools, on-demand functional annotation of
taxonomic features directly in R has remained challenging. To make this simple,
ariadne offers tools to rapidly annotate features with taxonomic or functional
modules based on online resources (BugSigDB, ChocoPhlAn, Web of Life and UniProt).
In addition, we provide tools to incorporate, analyse and visualise module
information into popular standardised containers, such as TreeSummarizedExperiment.


# Tutorial

## Installation

```{r install, eval=FALSE}
if( !requireNamespace("BiocManager", quietly = TRUE) ){
    install.packages("BiocManager")
}

BiocManager::install("ariadne")
```

## Explore the resource graph

ariadne combines different resources into a single graph highlighting the links
between biological entities. The resource graph can be imported as an igraph
object with the function `ariadne`.

```{r lib, message=FALSE, warning=FALSE}
library(ariadne)

# Import resource graph
graph <- ariadne()

# View graph
summary(graph)
```

The graph can also be visualised with `plotPath`.

```{r plot}
# Plot resource graph
plotPath(graph)
```

Let's say we need to find all links between gene functions and enzyme activity
represented by KEGG Orthologues (KO) and Enzyme Commission (EC) ids,
respectively. The function `searchPath` can be used to find the `k` shortest
paths in the graph.

```{r}
# Search for paths from ko to ec
searchPath(graph, ko ~ ec, k = 3)
```

```{r}
plotPath(graph, ko ~ ec, include = "uniref90", focus = TRUE)
```

We decide to go through the simplest path, which corresponds to a `k` of 1. The
links between KO and EC are found by the function `weavePath`, which fetches
the necessary resources and weaves a path from KO to EC.

```{r}
# Weave path from ko to ec
ko2ec <- weavePath(graph, ko ~ ec, k = 1)

# View some links
head(ko2ec)
```

The links are provided as a **linkmap**, a 2-column table where each row is a
pair of ids (one KO and one EC). When available, the feature names are returned
as a third optional column.

## Incorporate relational knowledge in a dataset

```{r}
library(mia)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Retrieve taxon names
tax.labs <- getTaxonomyLabels(tse, make.unique = FALSE)
tax.labs <- sub("^.+:", "", tax.labs)
```

```{r}
# butyrate
# addModules
# butyrate producer abundance
```


```{r}
tax2ec <- weavePath(
    graph,
    taxname ~ ec,
    include = "uniref90",
    init = tax.labs,
    factor = 8
)

head(tax2ec)
```

# Resources

## Citation

We hope that ariadne will be useful for your research. Please use the following
information to cite the package and the overall approach. Thank you!

```{r citation}
## Citation info
citation("ariadne")
```

## Background Knowledge

ariadne originates from the joint effort of the R/Bioconductor community. It is
mainly based on the following software:

- [R](https://www.r-project.org/), statistical programming language [@core2025r]
- [mia](https://bioconductor.org/packages/release/bioc/html/mia.html),
  framework for microbiome data analysis [@borman2025mia]
- [anansi](https://github.com/thomazbastiaanssen/anansi/),
  Annotation-based Analysis of Specific Interactions [@bastiaanssen2023anansi]
- [TreeSummarizedExperiment](https://bioconductor.org/packages/release/bioc/html/TreeSummarizedExperiment.htmlm), S4 container for hierarchical data [@huang2020treesummarizedexperiment]

## Help

You can reach us by one of the communication channels listed
[here](https://microbiome.github.io/OMA/docs/devel/pages/support.html). We are
happy to receive questions, suggestions as well as contributions. For the last
point, check the
[contributor guidelines](https://microbiome.github.io/mia/CONTRIBUTING.html).

# Reproducibility

R session information:

```{r reproduce, echo=FALSE}
# Session info
options(width = 120)
sessionInfo()
```
