---
title: "# ariadne: module integration by on-demand translation of annotations using R"
author: 
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
# bibliography: ../inst/REFERENCES.bib
date: "`r doc_date()`"
package: "`r pkg_ver('ariadne')`"
vignette: >
  %\VignetteIndexEntry{ariadne}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

ariadne provides a rapid and robust approach to incorporate taxonomic
and functional module analyses into your bioinformatic workflow. Despite the
presence of specialised bioinformatic tools, on-demand functional annotation of
taxonomic features directly in R has remained challenging. To make this simple,
ariadne offers tools to rapidly annotate features with taxonomic or functional
modules based on online resources (BugSigDB, ChocoPhlAn, Web of Life and UniProt).
In addition, we provide tools to incorporate, analyse and visualise module
information into popular standardised containers, such as TreeSummarizedExperiment.

## Motivation

# Tutorial

## Installation

```{r install, eval=FALSE}
if( !requireNamespace("BiocManager", quietly = TRUE) ){
    install.packages("BiocManager")
}

BiocManager::install("ariadne")
```

## Explore ariadne graph

```{r}
library(ariadne)

# 
graph <- ariadne()

#
plotPath(graph)
```

```{r}
searchPath(graph, taxname ~ ec, k = 10)
```

```{r}
plotPath(graph, taxname ~ ec, include = "uniref90", focus = TRUE)
```

```{r}
library(mia)

# Import dataset
data("Tengeler2020", package = "mia")
tse <- Tengeler2020

# Retrieve taxon names
tax.labs <- getTaxonomyLabels(tse, make.unique = FALSE)
tax.labs <- sub("^.+:", "", tax.labs)
```


```{r}
tax2ec <- weavePath(
    graph,
    taxname ~ ec,
    include = "uniref90",
    init = tax.labs[1:5],
    factor = 8
)

head(tax2ec)
```

# Resources

## Citation

We hope that ariadne will be useful for your research. Please use the following
information to cite the package and the overall approach. Thank you!

```{r citation}
## Citation info
citation("ariadne")
```

## Background Knowledge

ariadne originates from the joint effort of the R/Bioconductor community. It is
mainly based on the following software:

- [R](https://www.r-project.org/), statistical programming language [@core2025r]
- [mia](https://bioconductor.org/packages/release/bioc/html/mia.html),
  framework for microbiome data analysis [@borman2025mia]
- [anansi](https://github.com/thomazbastiaanssen/anansi/),
  Annotation-based Analysis of Specific Interactions [@bastiaanssen2023anansi]
- [TreeSummarizedExperiment](https://bioconductor.org/packages/release/bioc/html/TreeSummarizedExperiment.htmlm), S4 container for hierarchical data [@huang2020treesummarizedexperiment]

## Help

You can reach us by one of the communication channels listed
[here](https://microbiome.github.io/OMA/docs/devel/pages/support.html). We are
happy to receive questions, suggestions as well as contributions. For the last
point, check the
[contributor guidelines](https://microbiome.github.io/mia/CONTRIBUTING.html).

# Reproducibility

R session information:

```{r reproduce, echo=FALSE}
# Session info
options(width = 120)
sessionInfo()
```
