---
title: "ariadne: Integrate Knowledge in your Bioinformatic Analysis"
author:
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
# bibliography: ../inst/REFERENCES.bib
date: "`r doc_date()`"
package: "`r pkg_ver('ariadne')`"
vignette: >
  %\VignetteIndexEntry{ariadne: Integrate Knowledge into your Bioinformatic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

All domains of biology have to do with relational information. From genes to
organisms, each and every biological entity has some link to other entities. For
example, bugs are friends or foes with other bugs, have foods they like and
foods they hate (sometimes to death). All these links we call relational
information, which becomes knowledge once it is smartly combined with other
pieces and kept safe and open in a wonderfully organised place, usually a
database.

Building a database from scratch is no easy task. Fortunately, nowadays
databases for all kinds of relational information are available: UniProt for
proteins, Rhea for reactions, KEGG for pretty much anything, and many others,
which keep growing together with the field they support. As a scientist, having
access to such vast knowledge from past research can truly make us "stand on the
shoulders of giants".

In practice, though, it is not always easy (or even possible) to leverage such
knowledge in routine bioinformatic analyses due to three main challenges. First,
data is usually deposited somewhere online, so there must be a fast and reliable
method to retrieve and import it in the local environment. Second, data size
often exceeds the disk space of a typical laptop, which requires smart ways to
efficiently subset and store data in a lightweight, accessible format. Third,
unless you are lucky, data will need a tidy-up before it is compatible with your
dataset and other resources, necessitating a systematic approach to integrate
information from multiple resources.

## Motivation

Sometimes biology seems like a huge labyrinth with no simple paths between its
building blocks (genes, proteins, organisms, etc.). Even when the path to take
is clear, many obstacles may hide on the way: inaccessible resources, long query
times, incompatible formats, just to name a few. Crossing the labyrinth without
a strategy is a seriously boring and discouraging task.

Luckily, ariadne is here to help. By "Assembling Relational Information Across
the Database Network", ariadne provides a harmonised graph of interlinked
resources and a rich toolkit to navigate its underexplored paths. 

The harmonised graph of resources was built with a reproducible pipeline hosted
in the companion package `ariadne.db`, and it is openly available on [Zenodo](https://doi.org/10.5281/zenodo.18788725), including its previous
versions.


## Implementation

Before describing how ariadne functions, let us explain some important jargon:

- **feature type**: a distinct type of biological entity which includes a set of
  unique elements labelled with ids. For consistency, all feature types are
  written in lower case characters. Examples: `genes`, (KEGG genes id), `ko`
  (KEGG Orthologues), `ec` (Enzyme Commission number) or `uniref90` (90%
  similarity UniRef sequence clusters).
- **level**: a unique element belonging to a feature type. The term _feature_
  could be used as well, but it is not ideal due to its many closely related
  meanings. Examples: 563 (NCBI id for E. coli), C00246 (KEGG compound id for
  butyrate) or P0DTC2 (UniProtKB id for the SARS-CoV-2 spike protein).
- **linkmap**: 2-column data.frame where columns are the feature types and rows
  the links between them. This is how relational information is represented in
  ariadne. It is often named with the syntax `x2y`, where x and y are the names
  of the feature types in the first and second columns, respectively. Sometimes,
  there is also a third column with the names for the target feature type y.
  Examples: `tax2cpd` (linkmap between taxon names and compounds).
- **resource**: a database from where linkmaps between feature types can be
  retrieved. Examples: KEGG, Rhea or UniProt.
- **weave**: the operation of sequentially linking feature types to generate
  a new linkmap between two feature types that originally were not directly
  linked. This operation follows the syntax `x ~ y`, where x and y are the
  origin and target feature type to be linked, respectively. Examples:
  `ko ~ ec`, `disease ~ drug`, `genes ~ metacyc`.

Looking inside ariadne:

- The resource graph is represented as an `igraph` object, where edges are
  resources and nodes feature types. This is a natural choice for relational
  data and enables applying graph theory to select the preferred path from one
  feature type to another. It can be also easily expanded with user-defined
  resources.
- Resources are queried via fast `httr2` requests or using pre-existing APIs.
  Whenever possible, even faster SPARQL queries are used to fetch linkmaps.
- Linkmaps are automatically cached with `BiocFileCache` and efficiently stored
  and imported using the Apache parquet columnar format from `arrow`.
- The `MultiFactor` class, designed for optimised factor operations, takes care
  of generating the desired linkmap by sequentially weaving linkmaps along the
  path from the origin to the target feature type.


## Applications

ariadne provides a graph-based approach to integrate relational knowledge in
routine bioinformatic analyses. As such, it can be used to annotate any type of
biological data when related databases are available. This opens up potential
applications in various corners of bioinformatics, from knowledge-driven
feature annotation and stratification to goal-oriented feature selection and
prediction.


With microbes as features, ariadne allows to functionally annotate them using
ChocoPhlAn or WoL and to stratify them by Bug Signatures or Gut Modules. (ideas
for examples with metabolites, proteins, genes etc?)


ariadne is interoperable with the Bioconductor ecosystem. In particular, the SummarizedExperiment (SE) data container and its derived classes can be directly
annotated or stratified by relational information with ariadne.


# Tutorial

## Installation

```{r install, eval=FALSE}
if( !requireNamespace("BiocManager", quietly = TRUE) ){
    install.packages("BiocManager")
}

BiocManager::install("ariadne")
```

## Explore the resource graph

ariadne combines different resources into a single graph highlighting the links
between biological entities. The resource graph can be imported as an igraph
object with the function `ariadne`.

```{r lib, message=FALSE, warning=FALSE}
library(ariadne)
library(igraph)

# Import resource graph
graph <- ariadne()

# View graph
summary(graph)
```

The graph can be explored using all native functions from the igraph package.
For example, we can sort its nodes by degree, that is, the number of adjacent
edges:

```{r}
node_degrees <- graph |>
    degree(mode = "all") |>
    sort(decreasing = TRUE)

node_degrees <- node_degrees[node_degrees >= 10]

knitr::kable(node_degrees, col.names = "degree")
```

In addition, we can list all the nodes in the neighbourhood of a certain node:

```{r}
neighbors(graph, "ko", mode = "all")
```

The graph can also be visualised with `plotPath`.

```{r plot}
# Plot resource graph
plotPath(graph)
```

Let's say we need to find all links between gene functions and enzyme activity
represented by KEGG Orthologues (KO) and Enzyme Commission (EC) ids,
respectively. The function `searchPath` can be used to find the `k` shortest
paths in the graph.

```{r}
# Search for paths from ko to ec
searchPath(graph, ko ~ ec, k = 3)
```

```{r}
plotPath(graph, ko ~ ec, include = "uniref90", focus = TRUE)
```

We decide to go through the simplest path, which corresponds to a `k` of 1. The
links between KO and EC are found by the function `weavePath`, which fetches
the necessary resources and weaves a path from KO to EC.

```{r}
# Weave path from ko to ec
ko2ec <- weavePath(graph, ko ~ ec, k = 1)

# View some links
head(ko2ec)
```

The links are provided as a **linkmap**, a 2-column table where each row is a
pair of ids (one KO and one EC). When available, the feature names are returned
as a third optional column.

# Resources

## Citation

We hope that ariadne will be useful for your research. Please use the following
information to cite the package and the overall approach. Thank you!

```{r citation}
## Citation info
citation("ariadne")
```

## Background Knowledge

ariadne originates from the joint effort of the R/Bioconductor community. It is
mainly based on the following software:

- [R](https://www.r-project.org/), statistical programming language [@core2025r]
- [mia](https://bioconductor.org/packages/release/bioc/html/mia.html),
  framework for microbiome data analysis [@borman2025mia]
- [MultiFactor](https://github.com/thomazbastiaanssen/anansi/),
  Annotation-based Analysis of Specific Interactions [@bastiaanssen2023anansi]
- [TreeSummarizedExperiment](https://bioconductor.org/packages/release/bioc/html/TreeSummarizedExperiment.htmlm), S4 container for hierarchical data [@huang2020treesummarizedexperiment]

## Help

You can reach us by one of the communication channels listed
[here](https://microbiome.github.io/OMA/docs/devel/pages/support.html). We are
happy to receive questions, suggestions as well as contributions. For the last
point, check the
[contributor guidelines](https://microbiome.github.io/mia/CONTRIBUTING.html).

# Reproducibility

R session information:

```{r reproduce, echo=FALSE}
# Session info
options(width = 120)
sessionInfo()
```
