---
title: "Database"
author:
  - name: Giulio Benedetti
    affiliation: University of Turku
    email: giulio.benedetti@utu.fi
output:
    BiocStyle::html_document:
        fig_height: 7
        fig_width: 10
        toc: true
        toc_float: true
        toc_depth: 2
        number_sections: true
        self_contained: true
        code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('ariadne')`"
vignette: >
    %\VignetteIndexEntry{Database}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

The way that ariadne retrieves and combines relational data is informed by its
companion data package [ariadne.db](https://github.com/Minotau-R/ariadne.db),
which contains information about the feature types and mappings available from
a handful of popular databases. From here on, we will refer to those databases
as _resources_ to distinguish them from the ariadne database itself.

The relational information of each resource is described by a graph, where nodes
and edges represent feature types and mappings between them, respectively. In
ariadne.db, these graphs are stored in the Graph Modelling Language (GML), which
is a standard format for relational data.

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r lib, message=FALSE, warning=FALSE}
library(ariadne)
library(igraph)
library(dplyr)
library(ggplot2)
```

When `ariadne` is executed, the graph for each resource is retrieved from the
database release on [Zenodo](https://zenodo.org/records/18788726) and combined
into a single igraph object. Some resources have multiple versions, which can
be selected through the argument `versions`. This defaults to the latest version found in the database.

```{r}
# Import resource graph
graph <- ariadne()

# Plot resource graph
plotPath(graph)
```

```{r}
graph_df <- igraph::as_data_frame(graph, what = "both")
edge_df <- graph_df$edges
node_df <- graph_df$vertices
```


```{r}
df <- edge_df |>
    group_by(source) |>
    summarise(edges = n(), nodes = n_distinct(c(from, to)))

df
```

```{r}
for( resource in df$source ){
    
    res_df <- edge_df[edge_df$source == resource, c("from", "to")]
    res_graph <- graph_from_data_frame(res_df)
    
    p <- plotPath(res_graph) + ggtitle(resource)
    print(p)
}
```

# Reproducibility

R session information:

```{r reproduce, echo=FALSE}
# Session info
options(width = 120)
sessionInfo()
```

